---
title: "Field vs Lab Chlorine Accuracy"
author: "Laura Grasso"
date: "2023-10-31"
output: pdf_document
---
              
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = TRUE)

```

# Statistical analysis for assessing results difference between lab and field analysis for Total Chlorine

field SL1000 
lab method: SM 4500-Cl D




## data 

data from LIMS for the 2023 year.
sites are from the routine daily routes. 
data considered only when both field and lab have been collected 

```{r}
library(magrittr)
library(wqr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(fuzzyjoin)


start <-'2022-10-19'

end <- '2023-10-19'

param<- c("Field-Chlorine Residual Total", "Chlorine Residual Total")

sites2 <- c(1101, 1102, 1103, 1107, 1300, 1301, 1302, 1303, 1401, 1601, 1602, 1603, 
            1608, 1701, 1702, 1708, 1710, 1713, 1716, 1718, 1719, 1721, 2401, 2403,
            2501, 2502, 2600, 2601, 2602, 2603, 2706, 2712, 2713, 3800, 3801, 3803, 
            3901, 3904, 3906, 3907, 3913, 4001, 4005, 7101, 7204, 7207, 7301, 7302, 
            7401, 7502, 7601)



db <- read_LIMS(site = NULL, 
                
                parameter = param, 
                
                start_date = start, end_date = end, 
                
                sample_class = "Routine Daily") %>%
  
  filter(!is.na(result)) %>%
  
  separate('date_time', into = c("date", "time"), sep = ' ') %>%
  
  group_by(lims_number, site, date) %>%
  
  arrange(lims_number) 

  db

```

# data preparation for binning 

created columns: 

- difference between lab and field (pivot wider)
- percentage difference (lab as reference)
- binning referring to difference 


```{r}
#define tags 


tags <- c("[-1, -0.70)","[-0.70,-0.50)", "[-0.50, -0.20)", "[-0.20, -0.10)",
          "[-0.10, -0.05)","[-0.05, -0.02)", "[-0.02, 0)", 
          "[0-0.02)","[0.02-0.05)", "[0.05-0.10)", "[0.10-0.20)", 
          "[0.20-0.30)", "[0.30-0.50)","[0.50-1.0)")

labels <- tags


db1 <- pivot_wider(db, id_cols = c("site", "lims_number", "date", "temperature", "units"),
                   
                   values_from = result, 
                   
                   names_from = parameter) %>%
 
   filter(!is.na(`Field-Chlorine Residual Total`) & !is.na(`Chlorine Residual Total`)) %>%
  
  
   mutate(
     
     fieldtcl2 = `Field-Chlorine Residual Total`, 
     
     labtcl2 = `Chlorine Residual Total`,
     
     "difference" = (labtcl2 - fieldtcl2),
 
     "percentdiff" = ((fieldtcl2 - labtcl2)/labtcl2)*100 ) %>% 
  
select( -c(`Field-Chlorine Residual Total`, `Chlorine Residual Total`)) %>%
  
mutate(bins= cut(difference, 
                 
                 breaks=c(-1, -0.70, -0.50, -0.20, -0.10, -0.05, -0.02, 0,
                                        0.02, 0.05, 0.10, 0.20, 0.30, 0.50, 1.0), 
                 
                 include.lowest = TRUE,
                 
                 labels = tags))
       
       



#this does the same thing as the frequency of difference in the line below 

#grouptag <- summary(db1$bins)


```


# proportion difference neg/pos 

a negative difference means that field chlorine > lab chlorine (this is what we would expect since for the time gap between collection and lab analysis)
a positive difference means that lab chlorine > field chlorine



```{r}


pos <- nrow(db1[db1$difference>0,])
neg <- nrow(db1[db1$difference<0,])
equal <- nrow(db1[db1$difference ==0,])

tot <- pos + neg + equal 
counts <- as.data.frame(cbind(pos,neg, equal, tot))
counts
#normalizing 
posn <- pos/tot
negn <-  neg/tot
equaln <- equal/tot
countsn <- as.data.frame(cbind("pos" = posn,"neg" = negn, "equal" = equaln, tot))

df <- rbind("counts" = counts, "norm" = countsn)

df1 <- round(df, 2)
df1

```
graph 
```{r}


# A sample data frame
#df <- data.frame(
#  category = rep(c("A", "B", "C"), each = 3),
 # value = c("pos", "neg", "equal"),
 # norm = c(0.4, 0.3, 0.3, 0.5, 0.2, 0.3, 0.6, 0.1, 0.3))
#df


# Plot the normalized counts as stacked bars
#ggplot(df, aes(x = category, y = norm, fill = value)) +
#  geom_bar(stat = "identity", position = "stack")

```


#difference range binning between field and lab chlorine.

geom bar comparing counts of chlorine samples vs difference between lab and field. 

```{r}


plot <- ggplot(db1, aes(x= bins)) + 
  
  geom_bar(aes(y= after_stat(count))) + scale_y_continuous(limits=c(0,3000)) +
  
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -1) +


  xlab("difference lab vs field")  

 plot2 <- plot +  labs(title = "Values Difference Field vs Lab Chlorine") +
   
   theme(axis.text.x = element_text(angle = 60, hjust = 1)) 
 


 plot2
 
 


```
## normalized geom bar 

same analysis but normalized 

```{r}



total_count <- length(db1$lims_number)  #normalization for the next step
 
db1s <- db1 %>%
  group_by(bins) %>%
  # mutate(frequency = n()) %>%
  # ungroup() %>%
  summarize(norm = n() / total_count)


plot_n <-
  
  ggplot(db1s, aes(x= bins, y = norm)) + 
  
  geom_bar(stat = 'identity') +
  
  scale_y_continuous(label=scales::percent) + 
  
  xlab("difference lab vs field") + ylab("normalized percent") +

  
  geom_text(aes(label=scales::percent(norm), vjust = -1, angle = 25)) 
  
  
 plot2n <- plot_n +  labs(title = "Values Difference Field vs Lab Chlorine") +
   
   theme(axis.text.x = element_text(angle = 60, hjust = 1))
 


 plot2n
 
 # freq <- db1 %>% group_by(bins) %>% count((bins))   ## to review
 
 density_plot <- ggplot(db1, aes(y = bins, x = temperature)) +
  ggridges::geom_density_ridges() +
   scale_x_continuous(limits = c(0, 35)) +
  # geom_density(fill = "lightblue", alpha = 0.5) +
  xlab("difference lab vs field") +
  ylab("Density")

 
 density_plot
# Combine the two plots

 


```

## binning by the lab result 

```{r}


tags1 <- c("[0, 0.20]","[0.20, 0.50)", "[0.50, 0.70)", "[0.70, 1.0)",
          "[1.0, 1.20)","[1.20, 1.50)", "[1.50, 1.70)", 
          "[1.70, 1.90)","[1.90, 2.10)", "[2.10, 2.30)", "[2.30, 2.60)", 
          "[2.60, 2.90)", "[2.90-3.50)")

labels <- tags1


db3 <- pivot_wider(db, id_cols = c("site", "lims_number", "date", "temperature", "units"),
                   
                   values_from = result, 
                   
                   names_from = parameter) %>%
 
   filter(!is.na(`Field-Chlorine Residual Total`) & !is.na(`Chlorine Residual Total`)) %>%
  
  
   mutate(
     
     "fieldtcl2" = `Field-Chlorine Residual Total`, 
     
     "labtcl2" = `Chlorine Residual Total`,
     
     "difference" = (labtcl2 - fieldtcl2),
 
     "percentdiff" = ((fieldtcl2 - labtcl2)/labtcl2)*100 ) %>% 
  
select( -c(`Field-Chlorine Residual Total`, `Chlorine Residual Total`)) %>%
  
mutate(bins= cut(labtcl2,  
                 
                 breaks=c(0, 0.20, 0.50, 0.70, 1.0, 1.20, 1.50, 1.70, 1.90, 2.10,
                          2.30, 2.60, 2.90, 3.50),
                 
                 include.lowest = TRUE,
                 
                 labels = tags1)) 



# to check the range of labtcl2
#tt <- unique(db1$labtcl2) %>%
#  as_tibble() %>%
#  arrange(desc(value))

#summary(tt)

```

# difference vs temperature

```{r}
library(ggplot2)
# library(epiR)
library(survival)
library(scales)

plotres <- ggplot(db3, aes(x = temperature, y = difference)) +
  geom_point() +  # Add points for the scatterplot
  stat_smooth(method = "auto", se = FALSE) +  # Add a smoothing line with standard error bands turned off
  scale_y_continuous(limits = c(-0.6, 1.0), breaks = breaks_extended(30)) +
  scale_x_continuous(limits = c(0,30), breaks = breaks_extended(30))

plotres
# db4 <- db3%>% group_by(temperature)

plot_temp <- db3 %>% 
  mutate(
    temperature = factor(as.character(round(temperature)), levels = 1:35)
  ) %>%
  ggplot(aes(x = temperature, y = difference)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(limits = c(-0.6, 1.0), breaks = breaks_extended(30))
  # scale_x_continuous(limits = c(0,30), breaks = breaks_extended(30))
plot_temp

```



# Lab Chlorine 
lab chlorine binned vs scattered difference  ### to be normalized 

```{r}
library(scales)
plot_cl <- ggplot(db3, aes(x = bins, y = labtcl2)) +
  
  labs(title = "Lab Chlorine ranges vs difference between Field and Lab", 
       
       x = "Lab Chlorine Ranges", y = "Cl2 mg/l") +
  
 scale_y_continuous(breaks = breaks_extended(30)) +

 # geom_text(stat = 'count', aes(label=..count..), vjust = -1) +

  geom_boxplot(outlier.shape = NA) + 
    
  geom_jitter(mapping = aes(x = bins, y = difference), data = db3, stat = "identity", 
              
              position = "jitter", inherit.aes = TRUE, alpha = 0.3, color = "blue") +

  theme_bw() + theme(axis.text.x = element_text(angle = 60, hjust = 1))

plot_cl


```




```{r}

#temperature vs difference
#db1 is being used 
#  ggplot(db1, aes(x = temperature, y = difference)) + 
#  geom_point() +
#  xlim(10,30)

# to check the range of labtcl2
#tt <- unique(db1$temperature) %>%
# as_tibble() %>%
#arrange(desc(value))

#summary(tt)


```


 

```{r}
#temperature vs chlorine + difference
#db4 <- db3 %>% 
  
  
  

#plot_temp <- ggplot(db4, aes(x = temperature, y = labtcl2)) +
  
 # labs(title = "Lab Chlorine ranges vs difference between Field and Lab", 
       
  #     x = "temperature", y = "CL2 mg/l") +
  
 #scale_y_continuous(breaks = breaks_extended(30)) +
 # xlim(10,30) +

 # geom_text(stat = 'count', aes(label=..count..), vjust = -1) +

  #geom_boxplot(outlier.shape = NA) 
    
 # geom_jitter(mapping = aes(x = bins, y = difference), data = db3, stat = "identity", 
 #             
 #             position = "jitter", inherit.aes = TRUE, alpha = 0.3, color = "blue") +

  #theme_bw()

#plot_temp





```



multiple linear regression

```{r}

model <- lm(db1$difference~db1$labtcl2, data = db1)
print(summary(model))

# r-squared = 

```
```{r}
model1 <- lm(db1$difference~db1$temperature, data = db1)
print(summary(model1))
```

```{r}
model2 <- lm(db1$temperature~db1$labtcl2, data = db1)
print(summary(model2))
```



## pearson correlation 
correlation between lab chlorine and difference 
```{r}
library("ggpubr")
ggscatter(db1, x = "labtcl2", y = "difference",
          add = "reg.line", conf.int = TRUE,
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "lab chlorine", ylab = "difference")
```


## bias analysis with Bland-Altman plot 

A Bland-Altman plot provides visual insights into the differences and variability between the two methods. Here's how to interpret a Bland-Altman plot:

1. plot elements:

x-axis : The x-axis represents the average or mean of the measurements from the two methods.
y-axis: The y-axis represents the difference between the measurements from the two methods (Method 1 - Method 2).
Data points: Each data point on the plot represents an individual measurement.

2. dashed line: is drawn at the mean difference (bias) between the two methods.
Provides a reference for the overall bias. Ideally, you want this line to be close to zero, indicating minimal bias.

3. Spread of data:

   - The spread or dispersion of data points around the dashed line indicates the agreement between the two methods.
   - Wider spread suggests greater variability or disagreement between the methods.
   - The spread of data points is often used to calculate the limits of agreement.

4. Limits of Agreement (LOA):

   - The LOA are typically represented as two parallel blue lines above and below the mean difference line.
   - These lines are often set at ±1.96 times the standard deviation of the differences.
   - The LOA represent the range within which approximately 95% of the differences between the methods are expected to fall, assuming a normal distribution.
   - The narrower the LOA, the better the agreement between the methods.


```{r}
db3$diff <- db3$labtcl2 - db3$fieldtcl2
db3$mean <- (db3$labtcl2 + db3$fieldtcl2)/2


# Calculate the LOA
loa <- c(mean(db3$diff) - 1.96 * sd(db3$diff), mean(db3$diff) + 1.96 * sd(db3$diff))

# Create the Bland-Altman plot
p <- ggplot(db3, aes(x = mean, y = diff)) +
  geom_point() +
  geom_abline(intercept = mean(db3$diff), slope = 0, color = "red", linetype = "dashed") +
  geom_hline(yintercept = loa[1], color = "blue", linetype = "dotted") +
  geom_hline(yintercept = loa[2], color = "blue", linetype = "dotted") +
  labs(x = "Mean of Two Methods", y = "Difference Between Methods") +
  ggtitle("Bland-Altman Plot with LOA")

p


```


## concordance correlation coefficient 

```{r}

l#ibrary(epiR)
#library(survival)

# Load the epiR package
#library(epiR)

# Calculate CCC
#ccc_result <- concordance(db3$fieldtcl2, db3$labtcl2)

# Print the CCC result
#cat("Concordance Correlation Coefficient (CCC):", ccc_result$rho_c, "\n")




```




analysis by site 
```{r, fig.width = 10, fig.height = 7}
#  analysis to group by site and different values of chlorine 
site_info <- wqr::get_site_info() %>% 
  select(loc_id, service_area)

db5 <- db3 %>%
  left_join(site_info, by = c("site" = "loc_id")) %>% 
  
  # group_by(site) %>%

  mutate(site = as.character(site))

unique_sites <- unique(db3$site)

sites_groups <- cut(seq_along(unique_sites), breaks = 2, labels = c("Group 1", "Group 2"))


site_group_mapping <- setNames(sites_groups, unique_sites)

# Add a new column to your dataset indicating which group each site belongs to
db3$site_group <- site_group_mapping[db3$site]


# Your previous plot code with 'site_group' added to facet_wrap


  
  plot_site <- ggplot(db5, aes(x = site, y = difference)) +
    
    geom_boxplot(outlier.shape = NA) + 
    facet_wrap(vars(service_area), drop = TRUE, scales = "free_x", ncol = 2) +
  
  labs(title = "site vs difference", x = "sites", y = "Cl2 mg/l difference") 
  
 #scale_y_continuous(breaks = breaks_extended(30)) +

 # geom_text(stat = 'count', aes(label=..count..), vjust = -1) +
    
  #geom_jitter(mapping = aes(x = bins, y = difference), data = db3, stat = "identity", 
              
    #          position = "jitter", inherit.aes = TRUE, alpha = 0.3, color = "blue") +
#
 #theme_bw() + theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
     # facet_wrap(~cut(site, n = 2, labels = c("First Half", "Second Half")))

plot_site
  
  
  
```



## split sites in two groups to have a clearer view 









```{r}



# Assuming 'sites' is the column containing non-numeric site names

# Determine the number of unique sites
#unique_sites <- unique(db3$site)

# Split the unique sites into two groups
#sites_groups <- cut(seq_along(unique_sites), breaks = 2, labels = c("Group 1", "Group 2"))

# Create a named vector mapping each site to its corresponding group
#site_group_mapping <- setNames(sites_groups, unique_sites)




#db5 <- db3 %>%
  
 # group_by(site) %>%
 # summarise(result = mean(difference), .groups = "keep") %>%
 # mutate(site = as.factor(site)) %>%
 #cut(seq_along(unique(as.numeric(site))), breaks = 2, labels = c("Group 1", "Group 2"))
  



# Add a new column to your dataset indicating which group each site belongs to
#db3$site_group <- site_group_mapping[db3$site]

# plot_site <- ggplot(db5, aes(x = site, y = difference)) +
    
 #   geom_boxplot(outlier.shape = NA) + 
  
 # labs(title = "site vs difference", x = "sites", y = "Cl2 mg/l difference") 5

```






## analysis for instrument and time fidderence between collection and analysis time 



```{r}

start <-'2022-10-19'

end <- '2023-10-19'

param<- c("Field-Chlorine Residual Total", "Chlorine Residual Total")

sites2 <- c(1101, 1102, 1103, 1107, 1300, 1301, 1302, 1303, 1401, 1601, 1602, 1603, 
            1608, 1701, 1702, 1708, 1710, 1713, 1716, 1718, 1719, 1721, 2401, 2403,
            2501, 2502, 2600, 2601, 2602, 2603, 2706, 2712, 2713, 3800, 3801, 3803, 
            3901, 3904, 3906, 3907, 3913, 4001, 4005, 7101, 7204, 7207, 7301, 7302, 
            7401, 7502, 7601)



dbtr <- read_LIMS(site = NULL, 
                
                parameter = param, 
                
                start_date = start, end_date = end, 
                
                sample_class = "Routine Daily",
                
                select_additional = c("date_time_analyzed" = "ANALYZED_ON", "ANALYZED_BY")) %>%
  
  filter(!is.na(result)) %>%
  
  separate('date_time', into = c("date", "time"), sep = ' ') %>%
  
  separate('date_time_analyzed', into = c("date_an", "time_an"), sep = ' ') %>%
  
  # group_by(lims_number, site, date) %>%
  
  arrange(lims_number)


#read_LIMS(parameter = "Chlorine Residual Total", select_additional = c("date_time_analyzed" = "ANALYZED_ON", "ANALYZED_BY"))



```




```{r}

analysts <- dbtr %>% 
  select(lims_number, parameter, ANALYZED_BY) %>% 
  
  pivot_wider(names_from = parameter, values_from = ANALYZED_BY) %>% 
  
  rename(lab_analyst = `Chlorine Residual Total`, field_analyst =
           
           `Field-Chlorine Residual Total`)

dbtr1 <- pivot_wider(dbtr, id_cols = c("site", "lims_number", "date", "time" ,"temperature", "units",
                                       "ANALYZED_BY", "date_an", "time_an"),
                   
                   values_from = result, 
                   
                   names_from = parameter) %>%
 
  # filter(!is.na(`Field-Chlorine Residual Total`) & !is.na(`Chlorine Residual Total`)) %>%
  left_join(analysts, by = "lims_number") %>% 
  
   mutate(
     
     "fieldtcl2" = `Field-Chlorine Residual Total`, 
     
     "labtcl2" = `Chlorine Residual Total`,
     
     "difference" = (labtcl2 - fieldtcl2),
 
     "percentdiff" = ((fieldtcl2 - labtcl2)/labtcl2)*100 ) %>% 
  
select( -c(`Field-Chlorine Residual Total`, `Chlorine Residual Total`))



```



















































